---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)

```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

ANSWER HERE: From this data, I can see that crimes involving vehicles (auto) are some of the highest, including: Accident, Accident with Impound, Auto Stolen, and Theft from Auto. The only other crime that is consistently reporting high number is Theft. From a reporting standpoint, that is important for the community to know what our highest crime is so that they can take precautions. However, I think that the most important crimes, while not as common, would be the violent ones: Assault, Assault Shooting, Assault with Weapon, Sex Offense, and Homicide. These would be what the community is most afraid of, and would want updates regularly of when it happens.
I think the ones that would make summarizing this data most difficult are the ones where there are subcategories. For example, in January, there was 435 accidents. The next category states that there were 316 accidents with impound. It's important to know that these aren't totally separate occurrences, as all the accidents with impound are counted apart fo the total accidents, meaning there was only 119 accidents that were without impound. Having these in separate categories is an important distinction, but when sumarizing the data, could lead reporters to double count.

### Task 4: Calculating Averages

Add up the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime using the total number of months, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_total <- pg_crime |> mutate(total_incidents= Jan + Feb + Mar + Apr+ May+ Jun+ Jul+ Aug+ Sep+ Oct+ Nov+ Dec)
pg_crime_averages <- pg_crime_total |> mutate(average_incidents= total_incidents /12)
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

ANSWER HERE: The Accident incidents have the highest average per month of 481.3, while the lowest is the 5.8 average in Homicides. The Robbery Residential showed up as NA because there are a few data points missing for the months of March and October. I don't know how we would carry on calculating this? Would we remove those months from the total and averages calculations?  


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> mutate(june_change= (Jun- average_incidents)/average_incidents * 100)
summer_crime_averages <- summer_crime_averages |> mutate(july_change= (Jul- average_incidents)/average_incidents * 100)
summer_crime_averages <- summer_crime_averages |> mutate(august_change= (Aug- average_incidents)/average_incidents * 100)
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

ANSWER HERE: There is a lot of fluctuations in the percentage changes. I don't think there is remarkable patterns here, and I think this shows that the theory that crime is more likely to occur in the summer months isn't necessarily true. While there are some crimes where the percentage is 34% increase from the average, there is still a lot of percentage decreases.

### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE:
In Prince George's County, three crime categories, homicide, stolen automobiles, and thefts from automobiles, were consistently above average in their incident reports during the summer months of 2024, though they differed in the size of their increase. 

Homicide was 37% above average in the month of June, with 8 incident reports, and stayed above-average for the two following months. Stolen automobiles hit its peak of 11% above the average with 442 reports in June, while thefts from automobiles were 15% above average in July, with 499 incidents.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
