---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2026.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2026.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

ANSWER HERE: There are a few salaries that are at 0 dollars. All the other salaries range from ~60,000 to ~190,000. I'm assuming this is because these are volunteer positions/ just unpaid positions. But this will affect the median and average of this data, because the paid position are way higher.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean(SALARY))
wh_salaries |> summarize(median(SALARY))

```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean(SALARY),
                         median(SALARY),
                           min(SALARY),
                           max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

ANSWER HERE:
Since the mean is higher than the median, there is a right skew (the higher numbers are pulling more weight) which makes sense because most of the salaries are between 60,000-197000. The minimum is being pulled down by the 0 outliers. The range between the minimum and maximum tells me that there are a lot of variety in salaries within the white house 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

ANSWER HERE
I would want to look into the employees earning 0. What are their job descriptions/responsibilities. How did they become non compensated government employees? I would research their position titles and the history of the person. I would also look into what previous White House jobs had 0 salaries and if this is normal or just something happening under President Trump.
