---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
library(janitor)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv")
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |> mutate(
  violent_rate_2019 = (violent2019 / pop2019 * 1000),
  property_rate_2019 = (property2019 / pop2019 *1000),
  violent_rate_2020 = (violent2020 / pop2020 * 1000),
  property_rate_2020 = (property2020 / pop2020 * 1000))
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?

ANSWER HERE
The highest violent crimes in 2019 was in Baltimore, but in 2020, the rate increased from 18 to 23, and was higher in Ocean City. Ocean City was also the highest in Property rates for both 2019 to 2020, where the rates increased from 103 to 121. This could be because during the pandemic, the beach was someplace outdoors where people could go while still being able to social distance. There might be more property crimes because of the airbnb nature fo these trips.

### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> mutate(violent_change = (violent2020- violent2019/ violent2019 * 100),
    property_change = (property2020-property2019 / property2019 * 100))
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

ANSWER HERE  Only 13 violent crimes went up during 2020, the rest having significant decreases. Baltimore had significant increases, definitely an outlier.

### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

ANSWER HERE
While most Maryland cities saw crime decrease from 2019 to 2020, Ocean City had the highest crime rates in both violent and property crimes.

Baltimore had the highest violent crime rate in 2019, before the pandemic, and still saw an increase in its rate in 2020. However, Ocean City's violent crime rate rose above Baltimore's to 23%. Ocean City also had the highest property crime rate before the pandemic and saw that rate increase even further in 2020. The increase may have been created because Ocean City might have been attracting more visitors during the pandemic as an outdoor destination where people could vacation while still maintaining social distance. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?
Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.
Do the results change your best story idea from the previous assignment? Why or why not?

```{r}
md_cities_percentiles <- md_cities_rates |> mutate(percentile = ntile(violent_rate_2020, 100))
```

There are 10 cities above the 90th percentile.

```{r}
md_cities_rates |> summarize(IQR(violent_rate_2020))
```
This doesn't change my story because it doesn't explain why Ocean City has higher crime rates.The IQR means that 50% of Maryland cities had 4.2 crimes per 1,000 people. Ocean city has 23 crimes per 1,000 people, so it would still be in my story as the highest rate. I think I would add the IQR into my story to compare how high it is to the rest of the cities.